---
title: "Quantifying an educational effect"
author: "Flynn Hill"
date: 2018-09-13
output: blogdown::html_page
tags: ["regression", "statistics", "power analysis"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this post, I will run through a simple experiment simulation using only base R, and demonstrate a way to do power analysis by using this simulation. Say we have a cohort of 500 first-year students commencing their second session of study. We decide to offer a special tutoring service to some of the students who did not go well in their first session. To do this, we arrange students in ascending order of their averaged first session final marks, and go down the list calling students and offering tutoring services until 50 students have taken up the offer. At the end of the second session, we want to see if this made a difference. Here we will consider a simple situation where all relevant variables have been taken into account, and the effect is therefore consistent for all students - if only real life were so simple!

Although I am usually a devotee of the tidyverse, to keep things simple and focus on the procedure, I have used only base R in this post.

### Simulation
Start with the basic parameters and a latent variable we'll call aptitude, which for simplicity's sake will be the only thing (besides a bit of luck) that determines students' marks. As a latent variable, the most appropriate scale for this is a standard normal, i.e. a normal distribution centred at zero with a standard deviation of one.

```{r}
set.seed(42)
n_students <- 500
n_tutoring <- 50
meanmark <- 60.0
marks_sd <- 12.0


aptitude <- rnorm(n_students, 0.0, 1.0)
students <- data.frame(aptitude)

```



Generate a reasonable looking distribution of first-session marks, based on students' aptitude, plus or minus a little bit of luck. 

```{r}

luck <- 2

students$first_session_marks <- rnorm(n_students, students$aptitude*marks_sd + meanmark, luck)

#It is possible for some values to fall off the scale.
students$first_session_marks[students$first_session_marks > 100] <- 100
students$first_session_marks[students$first_session_marks < 0] <- 0


hist(students$first_session_marks, breaks = 15)

```

Now we start calling students to offer them tutoring, starting from those with the lowest first session marks. Here, the probability of a student accepting tutoring is 0.8 - first_session_mark/100, i.e. students with lower marks are more likely to accept tutoring.

```{r}

students <- students[order(students$first_session_marks),]

n_tutoring_places_accepted <- 0
tutored <- rep(0, n_students)
i <- 1
while (n_tutoring_places_accepted < n_tutoring & i < n_students) {
 rval <- runif(1) 
 if (rval < 0.8 - students$first_session_marks[i]/100){
   tutored[[i]] <- 1
   n_tutoring_places_accepted <- n_tutoring_places_accepted + 1
 }
 i <- i + 1
}

students$tutored <- tutored

sum(tutored)

```

Now simulate second session marks. These will also be based on students' underlying aptitude, and will be similar to the first session marks, but because the courses have changed there may be a different overall mean and scaling of marks with respect to to aptitude. To keep things simple (and unrealistic), students' underlying aptitude has not changed. In this model, there is an intercept term that gives the average (untutored) mark, a slope term that scales the mark with respect to aptitude, and another interept term for the effect associated with tutoring. Realised values of the second session mark are then drawn from Gaussian distributions centred at these calculated mark values, with a defined standard deviation that is the same for all students. Thus, for student $i$ we have:

$$\text{Second session mark}_i \sim \text{Normal}(\mu_i,\sigma)\\ \mu_i = \beta_0 + \beta_1\text{aptitude}_i + \beta_2\text{Tutored}_i $$
Define the parameters, and generate the second session marks according to the model. Here, we have very ambitiously assumed that tutoring increases students marks (to say nothing of whether it changes their aptitude) by 3 points.

```{r}
beta_0 <- 58 #This is the mean mark in second session for untutored students
beta_1 <- 0.9
beta_2 <- 3
sigma <- 2 #Luck in second session.

students$second_session_marks.mu <- beta_0 + beta_1*students$aptitude*marks_sd + beta_2*students$tutored
students$second_session_marks <- rnorm(n_students, students$second_session_marks.mu, sigma)

#Once again, it is possible that some values could have fallen off the scale, and this is one simple way to fix that:
students$second_session_marks[students$second_session_marks > 100] <- 100
students$second_session_marks[students$second_session_marks < 0] <- 0


hist(students$second_session_marks, breaks = 15)

```

Plot the first and second session marks, coloured by whether tutoring was accepted.

```{r}

plot(students$first_session_marks, students$second_session_marks, col = factor(students$tutored, levels = c(0,1)))

```

Students who received tutoring have had a small but appreciable bump in their marks. To measure the bump in marks, fit a linear regression, with a term for the tutoring effect. Students who were tutored have a value of 1 for the variable 'tutored', while those who weren't tutored have a value of 0, so any coefficient on this variable is essentially an intercept term.

```{r}
regression_model <- lm(second_session_marks ~ first_session_marks + tutored, data = students)
summary(regression_model)
```


\  

Note that we are now regressing against first session marks as a measure of the underlying aptitude from which the marks were drawn. Nonetheless, the model has retrieved fairly reasonable estimates for the coefficients - no surprise, they are a less noisy than real data, all relevant variables have been included, and the effect size is appreciable and consistent.


Now that we can see that the coefficient on 'tutored' is non-zero with p < 0.05, we can also do an analysis of variance to see whether the reduction in the residual sum of squares caused by including the tutored variable is significant. For the purpose of assessing whether tutoring has an effect, this is not the statistic of interest. It can be useful, however, in situations like when we want to make a predictive model, and include all the variables that will have a noticeable effect.

```{r}
anova(regression_model)
```

Indeed it is.

\  

An interesting point to note here is that some of the students who were offered tutoring may have received that offer because they were unlucky in first session and got marks a bit lower than what would be expected for their aptitude. In the second session, these students could be expected to get better marks regardless of being tutored - this is called [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean), and could be very problematic for a study such as this one.


\  

### Power analysis by simulation

A good idea before doing a study like this is to calculate its power - the probability of rejecting the null hypothesis when the null hypothesis is false (at a given significance level). This calculation can either help us to decide how many study subjects are needed to reach a desired power level, or if we are unable to increase the number of study subjects it can at least give us a realistic appreciation for the power of our study. There is a nice illustration of the concept of power in [this post](http://rpsychologist.com/d3/NHST/) by Kristoffer Magnusson.



We can use this simulation to find the power for an estimated effect size, or to find the minimum effect size needed to achieve a given power. By replicating the above simulation thousands of times, we can quantify the fraction for which the p-value of the tutoring effect coefficient was below the significance level. It is important to note that the effect size here is measured in units of marks, whereas effect sizes are often put into standard units by dividing by the standard deviation. The power simulation uses a simpler model to decide which students receive tutoring - for the bottom 100 students, a random 50 % accept tutoring. This mimics a situation in which we have no model for who is more likely to accept tutoring. The results of this power analysis could therefore mislead us.



```{r}

power_simulation <- function(n_tutored = 50, n_untutored = 450, tutoring_effect, sigma_secondsession, beta0 = 58, beta1 = 0.9, marks_sd = 12, method = "pval"){
  #Assign aptitude scores and generate first session marks
  nstudents <- n_tutored + n_untutored 
  aptitude <- rnorm(nstudents, 0, 1)
  students <- data.frame(aptitude)
  students$first_session <- rnorm(nstudents, students$aptitude*marks_sd + 60, 2)
  students <- students[order(students$first_session),]
  #Randomly assign tutoring to 50% of students in the bottom n_tutored*2 of first session marks.
  tutored.indices <- sample(seq(1,n_tutored*2), n_tutored)
  tutored <- rep(0, nstudents)
  tutored[tutored.indices] <- 1
  students$tutored <- tutored
  #Generate second session marks from aptitude, with a bump in marks for those who were tutored
  students$second_session <- rnorm(nstudents, beta0 + beta1*students$aptitude*marks_sd + tutoring_effect*students$tutored, sigma_secondsession)
  #Fit  a linear regression to get the p-value on the tutoring effect estimate.
  lmod <- summary(lm(second_session ~ first_session + tutored, data = students))
  if (method == "pval"){return(lmod$coefficients[3,4]) #The pvalue for the tutoring_effect estimate.
    #An option to return the coefficient value only if it is significant.
  } else if (method == "coeff" & lmod$coefficients[3,4] < 0.05){return(lmod$coefficients[3,1])
  } else if (method == "coeff" & lmod$coefficients[3,4] >= 0.05){return(NA)}
}

pvals <- replicate(10000, power_simulation(n_tutored = 50, 
                                           n_untutored = 450, 
                                           tutoring_effect = 1, 
                                           sigma_secondsession = 2))

mean(pvals < 0.05)


```


So, we have a calculated power of `r round(mean(pvals <= 0.05),1)` when there is an effect of +1 mark. The histogram below shows values of the coefficent for all simulations in which the p-value was below 0.05. Surprisingly, the true value of 1 is at the lower end of the range here. Selecting for only the statistically significant values has selected for those with inflated estimates, and the regression to the mean problem mentioned earlier is probably also making a contribution. In a noisier simulation, or one with a smaller effect size, the tutoring estimate could possibly even come out as both negative *and* statistically significant. 

```{r}
tutoring_effects <- replicate(10000, power_simulation(n_tutored = 50, 
                                                      n_untutored = 450, 
                                                      tutoring_effect = 1, 
                                                      sigma_secondsession = 2, 
                                                      method = "coeff"))

hist(tutoring_effects)
```


\ 

### Concluding remarks


Going into a power analysis it is common to have a personal bias to thinking our effect size will be greater than it really is, and that the data are less noisy than they really are. Without going too much into the whys and wherefores of power analysis and null hypothesis significance testing, [it is good practice](http://www.stat.columbia.edu/~gelman/research/published/PPS551642_REV2.pdf) to use a conservative estimate of the effect size based on external information (e.g. prior studies). If we wanted to make a proper attempt at doing a power analysis for an experiment like the one described above, we would need to have reasonable estimates for the distributions of marks, and a measure of how well correlated these are for students from session-to-session, to figure out the relationship to the underlying aptitude. If we go a step further and take into account that individual students' aptitudes can be on an upward or downard trajectory, it starts to get very complicated.



\  

